Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
small tweaks

In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore we
apply some transformation on the types to keep the size of types in check.

\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
    $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\  
\end{tabular}

Assuming that type sequences can be reordered freely, redundant types can be
deleted, types that are subsumed by subtyping can be removed, and nullable
types can be introduced to replace \null.

@@

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \contractr.
\begin{itemize}
\item Firstly, we faced the problem of type-checking arguments in a
  non-strict language while retaining the non-strict semantics. Recall that when a
  function is called in R, parameters are bound to unevaluated code thunks
  called promises~\cite{oopsla19}, instead of values which can be
  immediately type-checked. Naively forcing promises at function call to
  obtain a value for type-checking would violate the language semantics,
  leading to incorrect results. To preserve the original semantics,
  \contractr modifies the promise by wrapping the unevaluated promise
  expression in a call to its type-checker along with the context necessary
  to associate the promise to the function and its parameter (for failure
  messages). The contract checking happens when the promise is evaluated,
  either inside the function or many calls deep.  This works, but with one
  wrinkle. GNU R has a bytecode compiler that can sometimes optimize away
  promises; in those cases, \contractr receives values that can be
  immediately type-checked. Non-strict semantics dictate that any
  computation related to the argument should happen on the first use of the
  argument. So type-checking at this point and issuing an error message
  would violate the non-strict semantics; if this argument is never used, it
  would be incorrect to type-check it at the beginning of call. To make this
  work, \contractr mutates the function parameter binding to a promise that
  wraps the argument value in a call to the type-checker, as in the previous
  case.
\item Secondly, we faced the problem of type-checking function return
  value. In R, the result of a function call is the result of evaluating the
  last expression in the function body; so an explicit return call is absent
  from most function definitions. Thus, there could be many potential
  sub-expressions in the function body which need to be type-checked against
  the return type. To address this, \contractr registers the type-checker to
  be called on the return value through a function exit hook. This hook is
  executed in the function call's environment after it has executed. While
  this works, we have to contend with yet another wrinkle, R interpreter can
  perform longjumps in its C implementation, which causes active R function
  calls on the stack to be discarded. When they are discarded, their exit
  hooks are called, which in this case calls the registered
  type-checker. But, these functions are in the middle of an active
  computation, so they don't have a return value to type-check. \contractr
  deals with this problem by allocating a unique sentinel object which
  serves as the return value for calls that are discarded. The function call
  exit hook does not call the type-checker if this unique value happens to
  be the return value of the call.
\end{itemize}
%
%
\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers
for adding types to untyped package code. Hence, we have paid a lot of
attention towards usability in its design. We discuss three such usability
features.
\begin{itemize}

\item \contractr enables automatic type-checking on importing without any
  user intervention. Just running \code{library(contractr)} R code is enough
  to insert contracts, both, in packages pre-loaded in the user's workspace,
  and, in packages that will be loaded eventually. To achieve this
  automation, \contractr relies heavily on R's reflective and dynamic
  capabilities. On loading, \contractr scans all the pre-loaded packages in
  the user's workspace and inserts contracts in functions for which type
  signatures are available. For all other packages installed on user's
  machine but not yet loaded in the workspace, \contractr sets up package
  load hooks which are executed by R when those packages are loaded. The
  hook registered by \contractr insert contracts to those package's
  functions. Thus, developers and users do not have to call the \contractr
  API to enable type-checking for packages at any point of their programming
  workflow. Furthermore, \contractr automatically removes contracts from all
  the package functions and restores them to their original state when it is
  unloaded, an operation that is rarely performed, but useful in interactive
  settings.

\item \contractr enables package authors to supply type declarations for
  their package without requiring package code modifications. Type
  declarations can be written alongside a \code{@type} section inside
  \roxygen function comment blocks. \roxygen is an R package that enables
  authors to add documentation to R functions in the form of organized
  plaintext sections which are automatically exported to R's latex style
  custom documentation format.  \roxygen is a widely adopted package, used
  by over XXX R packages in CRAN. It provides an extension API for adding
  custom documentation tag sections in comment blocks. The sections are
  processed by \roxygen before package build step and a registered method
  for each custom tag section is invoked with the section data and
  documented object. \contractr uses this mechanism to register a hook for a
  custom \code{@type} tag. This hook parses the type declarations from the
  \code{@type} sections, extracts the function name from the documented
  function definition, and stores them in a TYPEDECLARATION file inside the
  package folder. This files is copied verbatim on package installation and
  picked up \contractr when the package is loaded. Here is an example of how
  a roxygen code block looks like with our type declaration:

@@

\item \contractr provides a very expressive API to the users, covering a
  variety of uses cases. During interactive development, developers can
  explicitly insert contracts by supplying the type declaration as a string
  argument to the \code{insert\_contract} API function. Conversely,
  contracts can be removed by calling the \code{remove\_contract} API on
  functions. Contracts can be selectively enabled or disabled for code
  blocks by wrapping them in calls to \code{ignore\_contracts} and
  \code{capture\_contracts} functions. This enables selective type-checking
  of code section during the development phase.  Furthermore, these
  functions return a data frame that contains all the information about
  failed and successful contract assertions in the wrapped code blocks. This
  is useful for post-hoc investigation. Multiple type-checking failures can
  introduce a lot of noise in the program output. To alleviate this problem,
  \code{contractr.severity} option can be set to \code{'silence'}.  This
  suppresses failure messages but \contractr still performs type-checking
  whose results can be explicitly obtained as a data frame from the
  \code{get\_contracts} API function. R suppresses printing of warnings on
  the terminal when a function issues too many warnings. This can hide
  type-checking failure messages. Setting \code{contractr.severity} flag to
  \code{'error'} turns the type-checking failure warnings to errors, which
  halts the program at the point of contract failure.

\end{itemize}

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & integer | null & 60578 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

@@

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
We were interested in knowing for each of the most common violations in Table~\ref{tbl:fail-breakdown-0}, how many different arguments had that type, and how many of those exhibited the contract failure in question.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by unique parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type \code{double[]} was passed to 15 different function parameters expecting a matrix, of which there are 1394 in total: 1.08\% of matrix-typed parameters were passed double vector values.

We see that even though the double vector and matrix issue was wildly prevalent in the raw, dynamic contract evaluation numbers, the number of actual function argument types that were violated is very small.
The story is similar with the double and integer coercion we mentioned earlier: it represents many dynamic contract failures, but very few of the integer vector-typed arguments have their contracts violated by double vector-typed values.
Rows 5 and 6 are interesting: we see that rather often arguments expecting time series data are passed double vector or matrix values.
This is a quirk of the \code{timeSeries} package, whose functions often accept matrices and vectors, converting them to time series in an ad hoc manner.
Note that the code coverage of the \code{timeSeries} tests, examples, and vignettes on the \code{timeSeries} package code is only 58.24\%, which is one possible explanation of why these contract failures are occurring: the types that \typetracer generates are only as good as the test code its run on.

@@

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
double[] & class<`matrix`> & 15 & 1394 & 1.08\\
double[] & integer[] & 44 & 657 & 6.70\\
character & class<`bignum`> | raw[] & 1 & 3 & 33.33\\
double & integer | null & 5 & 20 & 25.00\\
double[] & class<`timeSeries`> & 19 & 35 & 54.29\\
\addlinespace
class<matrix> & class<`timeSeries`> & 19 & 35 & 54.29\\
double[] & double & 100 & 5021 & 1.99\\
double & class<`data.frame`> & 1 & 900 & 0.11\\
double[] & class<`data.frame`> & 6 & 900 & 0.67\\
double & integer & 48 & 529 & 9.07\\
\bottomrule
\end{tabular}

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Results from Table~\ref{tbl:fail-breakdown-0}, broken down by occurrences of the expected type as a parameter type.}
 \end{table}

Table~\ref{tbl:fail-breakdown-2} presents data on the most frequently violated contracts amongst the most frequently occurring argument types.
We selected argument types which were in the 90th percentile of argument type occurrences, computed the most frequent type signature violations among them, and reported the most frequently violated contracts together with the type of the value that violated that contract. 
The first row of the table reads: 44 function arguments with integer vector type are passed double vectors instead, and 657 arguments have integer vector type.
Had we failed to capture some key usage pattern of R with our type annotation framework, we would likely see it here, and we can see this in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was obtained identically to Table~\ref{tbl:fail-breakdown-2} with a percentile of 80th percentile instead.
The most frequent argument type violation pattern in that of data frames being passed to arguments expecting tibbles: this occurs in only 22\% of such arguments, and represents cases where tests did not adequately cover all valid function inputs.
Separate from the issue of testing, we can capture this behaviour with user-defined subtyping, as data frames are subtypes of tibbles, but this necessitates a proper object-oriented type system for R which is outside the scope of this work.

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
double[] & integer[] & 44 & 657 & 6.70\\
\textasciicircum{}logical[] & null & 158 & 2756 & 5.73\\
character[] & character & 183 & 4102 & 4.46\\
character & null & 117 & 2756 & 4.25\\
double & null & 107 & 2756 & 3.88\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-2}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile.}
 \end{table}

\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
class<data.frame> & class<`data.frame`, `tbl`, `tbl\_df`> & 63 & 284 & 22.18\\
class<matrix> & class<`array`> & 13 & 70 & 18.57\\
\textasciicircum{}logical[] & logical | null & 9 & 49 & 18.37\\
character[] & list<character[]> & 20 & 109 & 18.35\\
complex[] & complex & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile.}
 \end{table}

In sum, we believe that this evaluation shows that the type signatures we generate from traces are quite good.
Only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed at runtime, representing failures in as few as \PROPARGSFAILINGASSERTS of argument types.
Even though \PROPFUNSFAILEDCHECKNOSTHREE of functions had at least one argument type involved in a failing contract check, these functions are under-tested, representing only \PERCCALLSBADFUNSINTESTS of calls observed while inferring types.

>>>>

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore we
apply some transformation on the types to keep the size of types in check.

\medskip
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
    $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\  
\end{tabular}
\medskip

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types.

@@

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R
has been a significant undertaking. We had to contend with the complex
design choices of R to make it work without surprises for the end-users. We
describe two key challenges that we faced in the development of \contractr.

Firstly, we faced the problem of type-checking arguments in a non-strict
language while retaining the non-strict semantics. Recall that when a
function is called in R, parameters are bound to unevaluated code thunks
called promises~\cite{oopsla19}, instead of values which can be immediately
type-checked. Naively forcing promises at function call to obtain a value
for type-checking would violate the language semantics, leading to incorrect
results. To preserve the original semantics, \contractr modifies the promise
by wrapping the unevaluated promise expression in a call to its type-checker
along with the context necessary to associate the promise to the function
and its parameter (for failure messages). The contract checking happens when
the promise is evaluated, either inside the function or many calls deep.
This works, but with one wrinkle. GNU R has a bytecode compiler that can
sometimes optimize away promises; in those cases, \contractr receives values
that can be immediately type-checked. Non-strict semantics dictate that any
computation related to the argument should happen on the first use of the
argument. So type-checking at this point and issuing an error message would
violate the non-strict semantics; if this argument is never used, it would
be incorrect to type-check it at the beginning of call. To make this work,
\contractr mutates the function parameter binding to a promise that wraps
the argument value in a call to the type-checker, as in the previous case.

Secondly, we faced the problem of type-checking function return value. In R,
the result of a function call is the result of evaluating the last
expression in the function body; so an explicit return call is absent from
most function definitions. Thus, there could be many potential
sub-expressions in the function body which need to be type-checked against
the return type. To address this, \contractr registers the type-checker to
be called on the return value through a function exit hook. This hook is
executed in the function call's environment after it has executed. While
this works, we have to contend with yet another wrinkle, R interpreter can
perform longjumps in its C implementation, which causes active R function
calls on the stack to be discarded. When they are discarded, their exit
hooks are called, which in this case calls the registered type-checker. But,
these functions are in the middle of an active computation, so they don't
have a return value to type-check. \contractr deals with this problem by
allocating a unique sentinel object which serves as the return value for
calls that are discarded. The function call exit hook does not call the
type-checker if this unique value happens to be the return value of the
call.

\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers
for adding types to untyped package code. Hence, we have paid a lot of
attention towards usability in its design. We discuss three such usability
features.

\contractr enables automatic type-checking on importing without any user
intervention. Just running \code{library(contractr)} R code is enough to
insert contracts, both, in packages pre-loaded in the user's workspace, and,
in packages that will be loaded eventually. To achieve this automation,
\contractr relies heavily on R's reflective and dynamic capabilities. On
loading, \contractr scans all the pre-loaded packages in the user's
workspace and inserts contracts in functions for which type signatures are
available. For all other packages installed on user's machine but not yet
loaded in the workspace, \contractr sets up package load hooks which are
executed by R when those packages are loaded. The hook registered by
\contractr insert contracts to those package's functions. Thus, developers
and users do not have to call the \contractr API to enable type-checking for
packages at any point of their programming workflow. Furthermore, \contractr
automatically removes contracts from all the package functions and restores
them to their original state when it is unloaded, an operation that is
rarely performed, but useful in interactive settings.

\contractr enables package authors to supply type declarations for their
package without requiring package code modifications. Type declarations can
be written alongside a \code{@type} section inside \roxygen function comment
blocks. \roxygen is an R package that enables authors to add documentation
to R functions in the form of organized plaintext sections which are
automatically exported to R's latex style custom documentation format.
\roxygen is a widely adopted package, used by over XXX R packages in
CRAN. It provides an extension API for adding custom documentation tag
sections in comment blocks. The sections are processed by \roxygen before
package build step and a registered method for each custom tag section is
invoked with the section data and documented object. \contractr uses this
mechanism to register a hook for a custom \code{@type} tag. This hook parses
the type declarations from the \code{@type} sections, extracts the function
name from the documented function definition, and stores them in a
TYPEDECLARATION file inside the package folder. This files is copied
verbatim on package installation and picked up \contractr when the package
is loaded. Here is an example of how a roxygen code block looks like with
our type declaration:

@@

\contractr provides a very expressive API to the users, covering a variety
of uses cases. During interactive development, developers can explicitly
insert contracts by supplying the type declaration as a string argument to
the \code{insert\_contract} API function. Conversely, contracts can be
removed by calling the \code{remove\_contract} API on functions. Contracts
can be selectively enabled or disabled for code blocks by wrapping them in
calls to \code{ignore\_contracts} and \code{capture\_contracts}
functions. This enables selective type-checking of code section during the
development phase.  Furthermore, these functions return a data frame that
contains all the information about failed and successful contract assertions
in the wrapped code blocks. This is useful for post-hoc
investigation. Multiple type-checking failures can introduce a lot of noise
in the program output. To alleviate this problem, \code{contractr.severity}
option can be set to \code{'silence'}.  This suppresses failure messages but
\contractr still performs type-checking whose results can be explicitly
obtained as a data frame from the \code{get\_contracts} API function. R
suppresses printing of warnings on the terminal when a function issues too
many warnings. This can hide type-checking failure messages. Setting
\code{contractr.severity} flag to \code{'error'} turns the type-checking
failure warnings to errors, which halts the program at the point of contract
failure.

@@

\begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & integer | null & 60570 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

@@

Turning our attention now to arguments, we found that only \PROPARGSFAILINGASSERTS of function parameters exhibited a failed assertion.
Table~\ref{tbl:fail-breakdown-0} showed the raw results from counting runtime occurrences, but that data alone does not tell the full story, as some failures may be overrepresented if e.g. a failing contract assertion was in a loop.
We were interested in a breakdown of type signature by parameter, and seeing how often those contracts were violated.
Table~\ref{tbl:fail-breakdown-1} breaks down the results from Table~\ref{tbl:fail-breakdown-0} by parameter position, folding away multiple identical failed contract assertions for the same parameter position.
The first row of this table reads: a value of type ${\bf chr}[]$ was passed to 183 different argument positions expecting a value of type ${\bf chr}$, representing 2.05\% of all arguments where a contract check failed.

The main takeaway from Table~\ref{tbl:fail-breakdown-1} is that there appear to be no pathological kinds of failing assertions: programmers are not disproportionately passing values of one type to arguments expecting values of another.
In other words, our type language appears to match with programmer intuitions: had we failed to account for a key, type-based assumption that R programmers make, we would see it at the top of the table.

@@

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Most common failing contract checks by type. Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
 \end{table}

%
%
% \subsubsection{Patterns in Failed Contract Assertions}

% While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.

% We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
% In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
% For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
% For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
% We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
